An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2019-11-19 manuscript/chapter_1.tex
Finished editing

@@

Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature \citep{}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were exlcuded from the initial pool. Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{}, leading to inaccurate estimates. Only plots with a stem density >10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.

@@

After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition. We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers. We set the threshold value to remove the top 5\% of plots with the highest nearest neighbour distances \citep{Otto}, removing \noutliers{} plots.

@@

We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{}. Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems forming a very small proportion of the total AGB in a plot \citep{}.

All stems >5 cm DBH were measured within each plot resulting in a total of 160,076 stems with measurements. A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual. For each stem, species, DBH and tree height to the top of the highest branch material were recorded. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which may cause varying measurement uncertainty by data source in the dataset. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from (Ryan, unpublished).

Aboveground biomass for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}. Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the mean of their respective genus.

\begin{equation}
	AGB_{est} = 0.0673 \times (\rho D^{2} H)^{0.976}
	\label{chave_agb}
\end{equation}

@@

Climatic data was collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}. Values of mean annual temperature (MAT) and mean annual precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature and precipitation seasonality were both calculated as the coefficient of variation of the daily data across the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), organic carbon density, and the percentage sand content of the soil. These data are a modelled product compiled from various remote sensed and directly measured data sources.

@@

\subsection{Data analysis}
Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness was preferred over raw Hill numbers as they are more interpretable, representing actual species numbers. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort across the region. Larger plots will tend to encompass more individuals, and therefore probabilistically more species \citep{}.

@@

Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot.

We calculated tree stand structural diversity for each plot by calculating the coefficient of variation of DBH and tree height.

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas with tree species diversity and relative abundance data. Groups were identified using unconstrained correspondence analysis and ordination. Plot data used in this study occurred in five vegetation type groups. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .

@@

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model indirect effects, which is impossible in multiple regression. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested by theory to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset. For example, moisture availability is known to affect productivity \citep{}, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration. Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.

Prior to model fitting, we specified a conceptual model with latent variables based on unmeasured factors known to affect AGB: moisture availability, soil fertility, tree species diversity and tree structural diversity (\autoref{con_mod}).

\begin{figure}[H]
\centering
	\input{include/path_diagram_concept}
	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural iversity, tree stem density, and AGB. Hypothesised paths of causal inference are depicted as arrows from predictor to response.}
	\label{con_mod}
\end{figure}

Observed variables were standardised to Z-scores for the SEMs. Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{}. Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan| \citep{}. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1. For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed. Precipitation seasonality, temperature seasonality, and mean temperature were also reversed in this way.

The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{}. While it is recommended by some to set exact factor loadings in the SEM from the regression coefficients of multiple regressions \citep{}, because some latent variables were regressed against both structural diversity and AGB, exact factor loadings from simple multiple regressions could not be used. We therefore allowed factor loadings to be estimated by the SEM itself.

First, we assessed the interacting factors of structural diversity and species diversity in determining AGB. We constructed a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity (see orange box \autoref{con_mod}). To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately. We compared path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between structural diversity, species diversity and AGB. Model fit was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We follow the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.95, TLI >0.95, RMSEA <0.06.

Second, we incorporated environmental covariates into our model to understand the relative effects of moisture availability and soil fertility on AGB both directly and indirectly via species diversity and stem density.

For models of the whole region which included all suitable plots, standardised path coefficients are reported in order to better compare the relative contribution of different paths within the model. When comparing the path coefficients of models using individual vegetation clusters, unstandardised path coefficients are used in order to more readily compare the same path coefficient across different models while accounting for variation in mean values between biogeographical clusters. Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.

To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create 900 datasets of similar size with varying median stem density. We used each of these datasets to fit the model including only tree species and structural diversity latent variables to predict AGB (see orange in \autoref{con_mod}). We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of the dataset.

We fitted separate moderation models to investigate whether there was an interaction effect whereby the strength of the relationship between species diversity and AGB was influenced by moisture availability and soil fertility. The \verb|lavaan| R package does not natively support moderation of latent variables in its model specification. Instead we manually calculated interaction variables for both soil fertility and moisture availability from the product of predicted values of these latent variables in a Confirmatory Factor Analysis (CFA). Interaction variables were the products: species diversity $\times$ moisture availability, and species diversity $\times$ soil fertility. These interaction variables were included as explanatory variables in multiple regressions alongside the latent variable of species diversity to predict AGB. Regression coefficients and model fit were analysed to determine the presence of interaction effects.

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.6\textwidth]{corr_mat}
	\caption{Correlogram of standardised observed variables used in the SEMs, with pearson correlation coefficients ($\rho$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation. Variables in bold and underlined on the axis labels were later reversed for SEMs to maintain positive correlations between observed variables within each latent variable. Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero. Colours of variable names group them into latent variables used in the SEMs: red = Soil fertility, blue = Moisture availability, green = species diversity, purple = structural diversity. See \autoref{corr_ci_tab} for a full table of correlation fit statistics.}
	\label{corr_mat}
\end{figure}

A simple set of correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB. Stem density had the strongest correlation with AGB of all variables ($\rho$ = 0.51, n = 1387, p <0.001). Environmental variables had weaker correlations with AGB than diversity variables. Only MAP ($\rho$ = 0.19, n = 1387, p <0.001), MAP seasonality ($\rho$ = -0.18, n = 1387, p <0.001 ), soil Organic C \% ($\rho$ = 0.19, n = 1387, p <0.001) and soil Sand \% ($\rho$ = -0.18, n = 1387, p <0.001) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero. (\autoref{corr_mat}).

As expected, there was a positive correlation between MAP and AGB, and a negative correlation between the seasonality of precipitation and AGB. MAT and the seasonality of temperature variation did not correlate with AGB. As expected, there was a negative correlation between soil sand content and AGB, and a positive correlation between soil organic Carbon AGB.

@@

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{struc_model_slopes}
	\caption{Standardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity. Due to all observed variables being standardised and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error. Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
	\label{struc_model_slopes}
\end{figure}

@@

When the structural diversity and species diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardised path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients. All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for clusters 3, 4 and 5 indicate a possible issue of low sample size for these clusters.

Cluster 5, which contained vegetation dominated by \textit{Colophospermum mopane} had no effect of tree species diversity on either AGB or stem density, in contast to the other vegetation clusters, but had a similar positive effect of structural diversity on AGB (\autoref{struc_model_slopes_all}).

The strongest total effect of tree species diversity on AGB was in cluster 3, which was relatively species poor compared to other vegetation clusters, dominated by \textit{Burkea africana}, \textit{Pterocarpus angolensis}, and \textit{Baikiaea plurijuga}, which are all large growing canopy tree species \citep{}.

The greatest effect of structural diversity was found in cluster 4, the most diverse vegetation cluster, with a median tree species richness of 11 species, dominated by the large growing \textit{Baikaea plurijuga}. Cluster 4 also had the highest median above-ground biomass and the highest median stem density (Table \autoref{clust_summ}). The effects of structural diversity on AGB were highly variable.

@@

We repeatedly sub-sampled the dataset of plots to build many datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density. The same SEM specification was used as above except that stem density was removed, incorporating the effects of species diversity and structural diversity on AGB (\autoref{con_mod}).  \autoref{sem_struc_stems_ha} shows that there is a peak in the total effect of tree species diversity on biomass at \textapprox{}700 stems ha\textsuperscript{-1}. There appears to be a minimum stem density threshold at \textapprox{}295 stems ha\textsuperscript{-1} below which there appears to be no effect of diversity on biomass. Interestingly, at high stem densities, the observed effect of species diversity on biomass begins to decline. The effect of structural diversity on biomass appears to decrease slightly with increasing stem density, though the relationship is less clear. At stem densities below \textapprox{}650 stems ha\textsuperscript{-1}  there is no observable effect of tree species diversity on structural diversity.

\begin{figure}[H]
\centering
	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves.}
	\label{sem_struc_stems_ha}
\end{figure}

@@

A model incorporating the latent variables of moisture availability and soil fertility showed that the effect of diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{path_diagram_full}). Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (\autoref{full_model_slopes}).

Moisture availability and soil fertility also had negligible direct effects on stem density, but both had a positive total effect on stem density via species diversity, where species diversity itself had a strong positive effect on stem density. As species diversity increases, stem density also increases.

Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was very weakly positive, while structural diversity itself had a positive effect on AGB. Again, stem density had a positive direct effect on AGB.

\begin{figure}[H]
\centering
	\input{include/path_diagram_full}
	\caption{Path diagram with regression coefficients for the full model incorporating environmental covariates and tree species diversity in all five clusters. Latent variables are shown as circles while observed variables are shown as rectangles. Path coefficients are solid arrows pointing from predictor to response. The observed variables which inform the unmeasured latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows. Measurement errors of exogenous variables are omitted for clarity.}
	\label{full_mod}
\end{figure}

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{full_model_slopes}
	\caption{Standardised path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots. Path coefficients are $\pm$ 1 standard error. Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
	\label{full_model_slopes}
\end{figure}

@@

Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass. Regression coefficient for the interaction effect of tree species diversity and moisture availability was significant ($\beta$ = 0.058, t(1383) = 2.562, p <0.05), while the interaction term of tree species diversity and soil fertility was not ($\beta$ = 0.034, t(1383) = 1.452, p = 0.147). As moisture availability increased, the effect of tree species diversity on AGB became stronger. At high tree species diversity, an increase in resource availability led to a larger increase in biomass.

\begin{figure}[H]
\centering
	\subfloat[]{
	\includegraphics[width=0.48\textwidth]{mois_int}
	}
	\subfloat[]{
	\includegraphics[width=0.48\textwidth]{soil_int}
	}
	\caption{Interaction plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b). All variables are standardised and centred to a mean of zero and a standard deviation of 1. Lines of best fit are drawn for standard deviations of moisture availability and soil fertility between -2 and +2. Moisture availability, tree species diversity and soil fertility are the values of latent variables the same as those used in the SEMs.}
	\label{int_plots}
\end{figure}

@@

The latent variable of tree species diversity had the strongest positive effect on woody above-ground biomass (AGB) in the environment+diversity SEM model (\autoref{full_mod}). Within the savanna woodlands of southern Africa we therefore find support for our hypothesis (H\textsubscript{1}), that higher tree species richness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{Liang2016, Cardinale2009}.

@@

\subsection{Future work}

>>>>

% STEM DENSITY

@@

Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup. Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature \citep{Ratnam2011, Hill2010}. Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were exlcuded from the initial pool. Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{}, leading to inaccurate estimates. Only plots with a stem density >10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014} (\autoref{data_clean_flow}).

@@

After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition. We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers. We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances \citep{Otto}, removing \noutliers{} plots.

@@

We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{Selaya2008}. Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems comprising a very small proportion of the total AGB in a plot \citep{}.

All stems >5 cm DBH were measured within each plot resulting in a total of 160,076 stems with measurements. A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual. For each stem we measured species, DBH and tree height to the top of the highest branch material. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from (Ryan, unpublished).

Woody AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}. Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}. Wood density for species without species level estimates was estimated from the mean of their respective genus.

\begin{equation}
	AGB = 0.0673 \times (\rho D^{2} H)^{0.976}
	\label{chave_agb}
\end{equation}

@@

Climatic data was collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}. Values of Mean Annual Temperature (MAT) and Mean Annual Precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature and precipitation seasonality were both calculated as the coefficient of variation of daily MAT and MAP, respectively, across the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org. C \%), and percentage soil sand content by volume (Sand \%). These data are a modelled product compiled from various remote sensed and directly measured data sources.

@@

\subsection{Data analysis}
Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness was preferred over raw Hill numbers as they are more interpretable, representing actual species numbers. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort. Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.

@@

Where $H'_{e}$ is an estimation of the Shannon diversity index of tree species by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot. We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH and tree height.

@@

Plots were assigned to vegetation type groups based on tree species composition. Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data. Groups were identified using unconstrained correspondence analysis and ordination. Plot data used in this study occurred in five vegetation type groups. See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .

@@

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package in R version 3.6.0 \citep{R2019}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}. A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is impossible in multiple regression. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset. For example, moisture availability is ecpected to affect AGB \citep{Saito2014, Campbell1996}, but moisture availability itself is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration. Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.

Prior to analysis, we specified a conceptual model with factors expected to affect AGB: moisture availability, soil fertility, tree species diversity, tree structural diversity and stem density (\autoref{con_mod}).

\begin{figure}[H]
\centering
	\input{include/path_diagram_concept}
	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural iversity, tree stem density, and AGB. Hypothesised paths of causation are depicted as arrows from predictor to response.}
	\label{con_mod}
\end{figure}

Observed variables were standardised to Z-scores prior to analysis. Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}. Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1. For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed for model fitting. Precipitation seasonality, temperature seasonality, and mean temperature were also reversed in this way to account for the direction of their effect on moisture availability.

The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{}. While it is recommended by some to set exact factor loadings in the SEM from the regression coefficients of multiple regressions \citep{}, because some latent variables were regressed against both structural diversity and AGB, exact factor loadings from simple multiple regressions could not be used. We therefore allowed factor loadings to be estimated by the SEM itself. We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, finding no significant difference between model specifications. Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.

First, we assessed the interacting factors of structural diversity and species diversity in determining AGB. We constructed a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity. To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately. We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB. Path coefficients show the effect of a path with other paths of inference held constant \citep{}. Model fit was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB. We follow the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.95, TLI >0.95, RMSEA <0.06.

To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create 900 datasets of similar size with varying median stem density. We used each of these datasets to fit the model including only tree species and structural diversity latent variables to predict AGB. We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled dataset.

Second, we incorporated environmental covariates into our model to understand the relative effects of moisture availability and soil fertility on AGB both directly and indirectly via species diversity and stem density. We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB.

We fitted separate moderation models to investigate whether there was an interaction effect whereby the strength of the relationship between species diversity and AGB was influenced by moisture availability and soil fertility. The \verb|lavaan| R package does not natively support moderation of latent variables in its model specification. Instead we manually calculated interaction variables for both soil fertility and moisture availability from the product of predicted values of these latent variables in a Confirmatory Factor Analysis (CFA). Interaction variables were the products: tree species diversity $\times$ moisture availability, and tree species diversity $\times$ soil fertility. These interaction terms were included as explanatory terms in multiple regressions alongside the latent variable of species diversity to predict AGB. Regression coefficients and model fit were analysed to determine the presence and form of interaction effects.

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.6\textwidth]{corr_mat}
	\caption{Correlogram of standardised observed variables used in the SEMs, with pearson correlation coefficients ($\rho$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation. Variables in bold and underlined on the axis labels were later reversed for SEMs to maintain positive correlations between observed variables within each latent variable. Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero. Colours of variable names group them into latent variables used in the SEMs: red = Soil fertility, blue = Moisture availability, green = tree species diversity, purple = tree structural diversity. See \autoref{corr_ci_tab} for a full table of correlation fit statistics.}
	\label{corr_mat}
\end{figure}

 Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB. Stem density had the strongest correlation with AGB of all variables ($\rho$ = 0.51, n = 1387, p <0.001). Environmental variables had weaker correlations with AGB than diversity variables. Only MAP ($\rho$ = 0.19, n = 1387, p <0.001), MAP seasonality ($\rho$ = -0.18, n = 1387, p <0.001 ), soil Organic C \% ($\rho$ = 0.19, n = 1387, p <0.001) and soil Sand \% ($\rho$ = -0.18, n = 1387, p <0.001) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero. (\autoref{corr_mat}).

As expected, there was a positive correlation between MAP and AGB, and a negative correlation between the seasonality of precipitation and AGB. MAT and the seasonality of temperature did not correlate significantly with AGB. As expected, there was a negative correlation between soil sand content and AGB, and a positive correlation between soil organic Carbon AGB.

@@

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{struc_model_slopes}
	\caption{Standardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity. Due to all observed variables being standardised and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error. Path coefficients where the standard error does not overlap zero are considered significant effects.}
	\label{struc_model_slopes}
\end{figure}

@@

When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardised path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients. All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for Clusters 3, 4 and 5 indicate a possible issue of low sample size for these clusters.

Cluster 5, which contained vegetation dominated by \textit{Colophospermum mopane} showed no effect of tree species diversity on either AGB or stem density, in contast to the other vegetation clusters, but had a similar positive effect of structural diversity on AGB (\autoref{struc_model_slopes_all}).

The strongest total effect of tree species diversity on AGB was in Cluster 3, which was relatively species poor compared to other vegetation clusters, dominated by \textit{Burkea africana}, \textit{Pterocarpus angolensis}, and \textit{Baikiaea plurijuga}, which are all large growing canopy tree species \citep{}.

The greatest effect of structural diversity was found in Cluster 4, the most diverse vegetation cluster, with a median tree species richness of 11 species, dominated by the large growing \textit{Baikaea plurijuga}. Cluster 4 also had the highest median above-ground biomass and the highest median stem density (Table \autoref{clust_summ}). The effects of structural diversity on AGB were highly variable.

@@

We repeatedly sub-sampled the dataset of plots to build many datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density. The same SEM specification was used as above (\autoref{struc_mod}) except that stem density was removed, incorporating the effects of species diversity and structural diversity on AGB (\autoref{con_mod}).  \autoref{sem_struc_stems_ha} shows that there is a peak in the total effect of tree species diversity on biomass at \textapprox{}700 stems ha\textsuperscript{-1}. There appears to be a minimum stem density threshold at \textapprox{}295 stems ha\textsuperscript{-1} below which there appears to be no effect of diversity on biomass. Interestingly, at high stem densities, the observed effect of species diversity on biomass begins to decline. The effect of structural diversity on biomass appears to decrease slightly with increasing stem density, though the relationship is less clear. At stem densities below \textapprox{}650 stems ha\textsuperscript{-1}  there is no observable effect of tree species diversity on structural diversity.

\begin{figure}[H]
\centering
	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves with standard error shaded bars.}
	\label{sem_struc_stems_ha}
\end{figure}

@@

A model incorporating the latent variables of moisture availability and soil fertility showed that the effect of diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{path_diagram_full}). Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}) (\autoref{full_model_slopes}).

Moisture availability and soil fertility also had negligible direct effects on stem density, but both had a positive total effect on stem density via species diversity, where species diversity itself had a strong positive effect on stem density (\rgid{}). As species diversity increases, stem density also increases.

Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was very weakly positive, while structural diversity itself had a positive effect on AGB (\rghb{}).

\begin{figure}[H]
\centering
	\input{include/path_diagram_full}
	\caption{Path diagram with regression coefficients for the SEM incorporating environmental covariates and tree species and structural diversity across all five vegetation type clusters. Latent variables are shown as circles while observed variables are shown as rectangles. Path coefficients are solid arrows pointing from predictor to response. The observed variables which inform the unmeasured latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows. Measurement errors of exogenous variables are omitted for clarity.}
	\label{full_mod}
\end{figure}

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{full_model_slopes}
	\caption{Standardised path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots. Path coefficients are $\pm$ 1 standard error. Path coefficients where the standard error does not overlap zero are considered significant effects.}
	\label{full_model_slopes}
\end{figure}

@@

Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass. The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant ($\beta$ = 0.058, t(1383) = 2.562, p <0.05), while the interaction term of tree species diversity and soil fertility was not ($\beta$ = 0.034, t(1383) = 1.452, p = 0.147). As moisture availability increased, the effect of tree species diversity on AGB became stronger.

\begin{figure}[H]
\centering
	\subfloat[]{
	\includegraphics[width=0.48\textwidth]{mois_int}
	}
	\subfloat[]{
	\includegraphics[width=0.48\textwidth]{soil_int}
	}
	\caption{Scatter plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b). All variables are standardised and centred to a mean of zero and a standard deviation of 1. Lines of best fit are drawn for standard deviations of moisture availability and soil fertility between -2 and +2. Moisture availability, tree species diversity and soil fertility are latent variables comprised of the same observed variable loadings.}
	\label{int_plots}
\end{figure}

@@

The latent variable of tree species diversity had a strong positive effect on woody above-ground biomass (AGB) in the environment+diversity SEM model (\autoref{full_mod}). Within the savanna woodlands of southern Africa we therefore find support for our hypothesis (H\textsubscript{1}), that higher tree species richness causes higher woody AGB. This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{Liang2016, Cardinale2009}.

@@

% \subsection{Future work}

% We hope that this study prompts further investigation of the relationship between woody productuivity and species richness and composition in southern African woodlands. While AGB is correlated with NPP,...
